Water discharge has a variable impact on sediment transport capacity in subglacial systems

### 2023-07-10 Delaney2023.tex
to ol

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\TODO}[1]{\textbf{TODO #1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian says:} \color{blue} #1} }
\newcommand{\leif}[1]{{\textbf{\color{red}Leif says:} \color{red} #1} }
\newcommand{\db}{\textit{DEBRIS}\,}
\newcommand{\cl}{\textit{CLEAN}\,}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
\newcommand{\unit}[1]{$\mathrm{#1}$}

@@

\ian{Alpine glacier vs Ice sheet}

@@

As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \cite<e.g.>[]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>[]{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \cite<e.g.>[]{muller1968,richards2003,ganti2016}.
Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.
Furthermore, thorough observations of water velocity and channel width in subglacial systems are limited and not contrasted with subaerial conditions at the glacier margins.

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher in the Swiss Alps (info) and the Leverett glacier on the Greenland Ice Sheet (info).
The hydrographs applied to the subglacial and subaerial parameterizations represent the subglacial flow of water as it leaves the glacier and moves into a subaerial flow in front of the glacier, as water moves through the catchment.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.

@@

@@

The water velocity and shear stress remain higher across the different friction factors and channel geometries examined in the ensemble of runs across averaging periods.
However, the Leverett case, some parameter combinations the result in greater variability in subaerial system compared to the subaerial channels.
The effect of channel evolving width largely counteracts the variability in water velocity, especially in the alpine glacier case.
Here the large diurnal cycles in the hydrograph can cause the maximum and minimum amounts of variability in subaerial total shear-stress to be greater than those in the subglacial system.
At the Leverett glacier, with reduced diurnal variability in water discharge (Figure\,\ref{fig:model_outs}), the variability in total shear-stress remains higher in the subglacial system compared to the subaerial one.
However, some variability in subaerial parameter combinations exceeds those subglacial ones.
Note however, that this occurs across the total of xxx parameter combinations examined.

@@

\section{Discussion}
\ian{add cautionary note as this is an example}

\ian{Zwally and tedstone effect with increased discharge... non-linear response of increasing water discharge. }

The relationship between water discharge variability and relative variability of the water velocity, shear stress, and width-integrated shear stress results from the fixed geometry in the glacier system and variable geometry in the subaerial system as they respond to changes in water discharge (Equations~\ref{eq:tau_t} and \ref{eq:v}).
Increased variability in sediment transport capacity with respect to discharge has several implications for sediment transport in subglacial systems compared to subaerial ones.
Shear stress is scaled to the $\frac{3}{2}$ power and relies on shear stress exceeding a threshold in sediment transport relationships  such as in \citeA{meyer1948}.
In the total sediment transport relationship \citeA{engelund1967}, the shear stress is scaled to $\frac{5}{2}$ power.
In both cases, the exponent greater than $1$ of shear stress in the sediment transport relationship magnifies the variability in sediment transport beyond the highly variable shear stress and width-integrated shear stress described above in subglacial environments (Figure~\ian{xxx}).

% Both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous \cite<Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}; e.g. >{yager2018}.
% \ian{Phillips and Jeromack 2016... stabilization here.}
% In the subaerial parameterization, this means that sediment is transported across all channel widths, as opposed to large flows such as floods when much sediment transport occurs \cite{wolman1960}.
% As a result, the  the subaerial systems variability in sediment transport capacity across the range of discharge values could be underestimated in the parameterization here (Figure~~\ian{xxx}).

@@

The parameterization here shows that the smallest variability in water discharge has the largest difference in relative variability in sediment transport metrics (Figure~~\ian{xxx}).
At locations such as the Greenland Ice Sheet, smaller diurnal variations in water discharge compared to many alpine catchments \cite<c.f.>{swift2005,riihimaki2005,vanas2017,hasholt2018}.
Therefore, the variability in sediment transport capacity in the transition from the ice sheet to the proglacial river could be more dramatic in Greenland than in alpine systems as water leaves the glacier and moves downstream.
Thus the disparity in sediment transport capacity between the subglacial and subaerial systems could be most pronounced in larger catchments, potentially resulting in temporally mismatched periods between when glaciers introduce sediment to these river systems  and when the river mobilizes it (Figures~\ref{fig:model_outs}).

@@

\ian{include comments about the limits of hysterisis analysis from glacial systems}

Large variations in sediment transport capacity below glaciers may result in sediment mobilization and deposition processes in close spatial and temporal proximity to each other below the glacier \cite{gimbert2016,perolo2018}.
The rapid increase and decreases in sediment transport capacity, for example over a diurnal cycle, in a subglacial system may compound the variability in sediment transport already present in subaerial systems \cite{williams1989,jerolmack2010}.
As a result, variations in sediment transport capacity could be responsible for the fluctuations in sediment discharge (``flushings'') from glaciers that are not attributed to variations in water discharge \cite<e.g.>[]{richards2003,swift2021}.
Aside from the sporadic nature of these events, the observed ``flushings'' could result from increases in water velocity resulting from reduced channel size, not water discharge, which increase sediment transport capacity.

@@

Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \cite<e.g.>[]{koppes2009a,willenbring2016,mariotti2021}.
The variability in subglacial sediment transport capacity presented here applies most generally to short-time scales responsible for the size of subglacial channels.
Yet, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of climatic perturbation compared to subaerial systems \cite{tofelde2021}.
This higher threshold of climatic perturbation may persist in addition to the stochastic nature of erosion and deposition in fluvial environments \cite{castletort2003,jerolmack2010,romans2016}.
For instance, sediment discharge records from two glaciers in the Swiss Alps show that $40$ to $50$\% of a season's sediment discharge occurs when water discharge is below the $75^{\mathrm{th}}$ quantile of the season \cite{delaney2018}.
Such a large quantity of sediment transported at lower water discharges may occur in part because sediment transport capacity responds to the size of the glacier conduit, in addition to water discharge (Section~\ref{sect:sub_mode}).

>>>>

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\TODO}[1]{\textbf{TODO #1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian says:} \color{blue} #1} }
\newcommand{\leif}[1]{{\textbf{\color{red}Leif says:} \color{red} #1} }
\newcommand{\alpine}{\textit{ALPINE}\,}
\newcommand{\icesheet}{\textit{ICESHEET}\,}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
\newcommand{\unit}[1]{$\mathrm{#1}$}

@@

@@

As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \cite<e.g.>[]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
However, some suggest that subglacial sediment discharge  $Q_s \propto Q_w^{\frac{9}{2}}$, assuming that water discharge, changing channel width and depth are each proportionally responsible for the variance of water velocity and sediment transport capacity \cite{alley1997,swift2021}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>[]{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \cite<e.g.>[]{muller1968,richards2003,ganti2016}.
Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.
Furthermore, thorough observations of water velocity and channel width in subglacial systems are limited and not contrasted with subaerial conditions at the glacier margins.

@@

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher in the Swiss Alps (info) and the Leverett glacier on the Greenland Ice Sheet (info).
The hydrographs applied to the subglacial and subaerial parameterizations represent the subglacial flow of water as it leaves the glacier and moves into a subaerial flow in front of the glacier, as water moves through the catchment.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, the Fieschergletscher case is referred to as \alpine{} and is exemplified by relatively thin ice thickness ($h_o$: xxx), low discharge (< $20$\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in discharge.
The Leverett glacier case is referred to as \icesheet{} and is exemplified by thick ice  ($h_o$: xxx), high water discharge (< $300$\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in discharge.

@@

The water velocity and shear stress remain higher across the different friction factors and channel geometries examined in the ensemble of runs across averaging periods.
However, the Leverett case, some parameter combinations the result in greater variability in subaerial system compared to the subaerial channels.
The effect of channel evolving width largely counteracts the variability in water velocity, especially in the alpine glacier case.
Here the large diurnal cycles in the hydrograph can cause the maximum and minimum amounts of variability in subaerial total shear-stress to be greater than those in the subglacial system.
At the Leverett glacier, with reduced diurnal variability in water discharge (Figure\,\ref{fig:model_outs}), the variability in total shear-stress remains higher in the subglacial system compared to the subaerial one.
However, some variability in subaerial parameter combinations exceeds those subglacial ones.
Note however, that this occurs across the total of xxx parameter combinations examined.
In both glaciers and across all variables, the variability reduces from one day to five day aggradations (Figure\,\ref{fig:multi_run}).

@@

\section{Discussion}

The relationship between water discharge variability and relative variability of the water velocity, shear stress, and width-integrated shear stress results from the fixed geometry in the glacier system and variable geometry in the subaerial system as they respond to changes in water discharge (Equations~\ref{eq:tau_t} and \ref{eq:v}).
Increased variability in sediment transport capacity with respect to discharge has several implications for sediment transport in subglacial systems compared to subaerial ones.
Shear stress is scaled to the $\frac{3}{2}$ power and relies on shear stress exceeding a threshold in sediment transport relationships  such as in \citeA{meyer1948}.
In the total sediment transport relationship \citeA{engelund1967}, the shear stress is scaled to $\frac{5}{2}$ power.
In both cases, the exponent greater than $1$ of shear stress in the sediment transport relationship magnifies the variability in sediment transport beyond the highly variable shear stress and total shear stress described above in subglacial environments (Figure~\ref{fig:Qw_vari}.

Both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous \cite<Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}; e.g. >{yager2018}.
In the subaerial parameterization, this means that sediment is transported across all channel widths, as opposed to large flows such as floods when much sediment transport occurs \cite{wolman1960}.
As a result, the  the subaerial systems variability in sediment transport capacity across the range of discharge values could be underestimated in the parameterization here (Figure~\ref{fig:Qw_vari}).
Furthermore, the similar variance of total shear stress shows (Figure~\ref{fig:multi_run}) that subglacial channels contain self-organizing feedbacks that limit the mobilization of sediment, as in fluvial systems \cite{phillips2016}.

@@

@@

The pressurized nature of subglacial water means that the relationship between water discharge and suspended sediment concentration, or sediment discharge could be disparate, especially over the timescales examined here.
The rapid increase and decreases in sediment transport capacity, for example over a diurnal cycle, in a subglacial system may compound the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
Large variations in sediment transport capacity below glaciers may result in sediment mobilization and deposition processes in close spatial and temporal proximity to each other below the glacier \cite{gimbert2016,perolo2018}.
As a result, variations in sediment transport capacity could be responsible for the fluctuations in sediment discharge (``flushings'') from glaciers that are not attributed to variations in water discharge \cite<e.g.>[]{richards2003,swift2021}.
Aside from the sporadic nature of these events, the observed ``flushings'' could result from increases in water velocity resulting from reduced channel size, not water discharge, which increase sediment transport capacity (Figure~\ref{fig:Qw_vari}).

@@

Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \cite<e.g.>[]{koppes2009a,willenbring2016,mariotti2021}.
The variability in subglacial sediment transport capacity presented here applies most generally to short-time scales responsible for the size of subglacial channels (Figure~\ref{fig:multi_run}).
Yet, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of climatic perturbation compared to subaerial systems \cite{tofelde2021}.
This higher threshold of climatic perturbation may persist in addition to the stochastic nature of erosion and deposition in fluvial environments \cite{castletort2003,jerolmack2010,romans2016}.
Results here suggest that increased variability in sediment transport in subglacial systems  could be particularly pronouced on ice sheets with large catchments and reduced variability.
For instance, sediment discharge records from two glaciers in the Swiss Alps show that $40$ to $50$\% of a season's sediment discharge occurs when water discharge is below the $75^{\mathrm{th}}$ quantile of the season \cite{delaney2018}.
Such a large quantity of sediment transported at lower water discharges may occur in part because sediment transport capacity responds to the size of the glacier conduit, in addition to water discharge (Figure~\ref{fig:model_outs} and Section~\ref{sect:sub_mode}).
